`respR` - An R package for the analysis of respirometry data

### 2017-11-17 manuscript/LaTeX/manuscript.Rmd
manuscript updates

4. Although designed specifically with aquatic respirometry in mind, the object-oriented approach of the package and the dimensionless nature of its analytical functions mean that parts of the package can easily be used to estimate any other linear relationships in other research disciplines."

keywords: "R, software, respirometry, metabolic rate"
thanks: "Replication files are available on the first author's Github account (http://github.com/januarharianto). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: nicholas.carey@sams.ac.uk."

@@

opts_chunk$set(collapse = TRUE, comment = "#>", tidy = TRUE, highlight = TRUE)
```

# Introduction
Metabolic rate is a fundamental trait associated with virtually all biological functions and is key in predicting patterns in ecology and conservation biology, from populations [@seibel_rate_2007; @barneche_scaling_2014] to ecosystems [@brown_toward_2004]. It has been widely investigated in studies of the effects of external stressors on organisms, and is increasingly used to study resilience in climate change-related studies where warming is expected to drive increased metabolism in ectotherms [e.g. @portner_climate_2002; @mcelroy_respiratory_2012; @carey_sea_2016; @delorme_effects_2016]. By far the most common metric used to assess physiological performance in animals, metabolic rate has been determined from whole organisms to the level of cells and tissues [@white_determinants_2013]. Metabolic rate is typically quantified using respirometry, which measures oxygen uptake over time [@lighton_measuring_2008].

There are four broad methodological approaches in respirometry: closed, intermittent flow, flow-through and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow mixing of the environment within the chamber. Oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from the $O_2$ timeseries by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$VO_2 = \Delta O_2V$$ where $\Delta O_2$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container [@lighton_measuring_2008].

In intermittent flow respirometry, $O_2$ concentration is measured as described above, but periodically the chamber is flushed with new water or air, returning it to initial conditions, resealed, and the experiment repeated [@svendsen_design_2016]. This technique is essentially closed respirometry, but with the incorporation of repeated measures. Depending on the metabolic rate metric being investigated, final respiration rate can be calculated as the mean of the measures [e.g. @carey_sea_2016], or the lowest or highest rates recorded in any trial [e.g. @stoffels_physiological_2015].

Flow-through respirometry involves a closed chamber, but with a regulated flow of air or water through it at a precisely determined rate. After equilibrium has been achieved, the oxygen concentration differential between the incurrent and excurrent channels, along with the flow rate, allows calculation of the oxygen extracted from the flow volume per unit time: $$VO_2 = (C_iO_2 - C_eO_2)FR$$ where $VO_2$ is the rate of $O_2$ consumption over time, $C_iO_2$ and $C_eO_2$ are the incurrent and excurrent $O_2$ concentrations, and $FR$ is the flow rate of water through the system [@lighton_measuring_2008].

A final method is open respirometry, in which an open tank or semi-enclosed area is used, but the input or mixing rate of oxygen from the surroundings is known or found to be negligible relative to oxygen consumption of the specimens [@leclercq_measurement_1999]. It is seldom used, but for some applications it is a sufficient and practical methodology [@gamble_open-top_2014]. The common equation used for open respirometry is: $$VO_2 = \Delta O_2V + \phi_d$$ where $\Delta O_2V$ is the slope of the regression that relates $O_2$ concentration to time, $V$ is the volume of the container and $\phi_d$ is the oxygen flux as determined by Fick's Law [@leclercq_measurement_1999].

@@

Respirometry studies are increasingly outputting large, high-resolution datasets and conducted over longer periods [e.g. 20 h ; @norin_intraspecific_2012]. In most cases, processing the data involves an *ad hoc* selection of data points with poorly reported criteria, and subsequent manual processing of the data subset(s) using a spreadsheet program (e.g. Microsoft Excel) or an integrated development environment (IDE, e.g. R or Matlab). These approaches can be tedious and time consuming especially when spreadsheet programs struggle with the vast datasets that are generated, while IDEs require a degree of expertise to use and have substantial learning curves. Dedicated software are also available to perform metabolic rate analyses, but many have costly licensing requirements (e.g. AutoResp by Loligo Systems), complicating or preventing their use on multiple machines, and are proprietary and closed-source, hindering scientific reproducibility and transparency.

A number of open-source R software packages have recently become available which are designed for, or are suitable to analyse respirometry data. The `respirometry` package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The `rMR` package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). The package `LoLinR` [@olito_estimating_2017], while not specifically coded for respirometry, provides a statistically robust method of detecting a "best-fit" regression, and performs very well at identifying truly linear subsets of a data series. Importantly, all of these packages assume that data are monotonic with evenly-spaced sampling periods with respect to time, an ideal structure but one where respirometry data often fails to meet in reality. They also do not have the tools to analyse complicated datasets to automatically determine useful metric such as $MO_{2,max}$ and $MO_{2,min}$.

Here we describe `respR`, an open-source R package designed to provide an efficient and reproducible workflow for the analysis of respirometry data (Table 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) determine volume, mass, and surface area specific oxygen uptake rates, (3) automatically detect maximal and minimal rates, $P_{crit}$, and (4) detect the most linear sections of data using traditional rolling regression and kernel density estimation techniques. The package facilitates the output as supplementary material of all stages of respirometry analyses; raw data, subsetting criteria, background corrections, units of measure, and resulting metabolic rates. This streamlines the reporting of analytical methods and allows reviewers or other investigators to fully reproduce or scrutinise the results. Other functions useful in processing respirometry data are avaliable and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). The utility of the `respR` package is shown here using  data collected for closed, intermittent and flow-through respirometry. We show how `respR` might be used to determine $MO_{2,max}$, $MO_{2,min}$, and $P_{crit}$.

@@

The `respR` package streamlines the analysis of respirometry data by partitioning the data processing workflow into several independent actions: (1) initially, data are imported, checked for common errors, plotted for visual inspection and exported as an object; (2) the object is then analysed to estimate the desired rate parameter; (3) diagnostic plots are automatically generated and the data can be further explored using the generic S3 `print()`, `plot()`, and `summary()` commands to verify the results; (4) adjustments to the data, either from background respiration and/or oxygen flux in open respirometry, are applied; (5) volume or mass-specific rate conversions, specific to the respiratory chamber and specimen(s), are calculated where necessary (Fig. 1). By design, each step of the workflow requires minimal manual manipulation of the data, and the user can pass the output of each function to the next. This object-oriented approach makes `respR` easy to use, even for novice users of R, and allows us (or anyone interested in the open-source project) to easily extend the functionality of the package, either by adding additional intermediate actions or substituting current actions with other code without breaking the rest of the package. Note, we use $RO_2$ in the following text to represent the dimensionless rate of $O_2$ use with regard to the time unit, and to distinguish it from other commonly reported metrics (e.g. $MO_2$, $VO_2$).

@@

Data should be formatted correctly before use in `respR`. The function `inspect.data()` extracts a two-column data frame from a multi-column dataset and performs error checks on a data frame of any size, while plotting the data for quick data visualisation. All data should be numeric; respR at this time has limited support for date/time data (e.g. `POSIXct` and `POSIXt` classes). Parsing  different time formats is extremely complex and beyond the scope of this package, particularly because such packages already exist (e.g. `lubridate`). Time data, being a continuous variable, should be sequential, without duplicates. Ideally, sampling frequency will be monotonic, but in many cases this may not be the case, such as experiments with irregular sampling intervals, or because machine drift and/or data dropouts frequently occur during regular sampling. Therefore, `respR` uses time-based data extraction methods by default in most of its analytical functions (except `pcrit()`), and correctly handles irregular timeseries data.

Using `inspect.data()` is optional -- the main functions in our package will readily work with data frames that were not processed by the function. Running `inspect.data()` is an exploratory step that flags potential issues before analysis. Error check results are summarised in the R console and all tests are saved in the output object with the locations (i.e. row numbers) of the errors:

@@

Background oxygen consumption rate, $RO_{2(bg)}$, is often measured to account for the contribution of microbial respiration, and applied as a correction to experimental rates [@rogers_new_2016]. Background rates typically account for a small percentage of experimental rates, and multiple “blank” experiments are usually conducted and the rates averaged to obtain a more accurate estimate of the correction [e.g. @carey_size_2014; @daoud_temperature_2007]. The function `calc.rate.bg()` uses simple linear regression to simultaneously process multiple background rate measurements. Data segments can be truncated before analysis by time period or row numbers. The results are stored as an object for later use.

@@

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ [e.g. @gordon_energetics_1989; @chu_ecophysiological_2017]. The function `calc.rate()` can manually extract and process data subset from respiromety data. Data subsets can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accommodate most, if not all, manual data selection requirements and allow for consistent reporting of methods and results. Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results.

For more complex respirometry parameters, the function `auto.rate()` uses novel rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs regressions of sample size $n$ iteratively across the length of the data (i.e. rolling regressions), one sample at a time (**Fig. 2a**). Thus, a total of $(m - n) + 1$ number of overlapping regressions are fitted, which are then ranked to obtain maximum and minimum values. If an interval-based regression is required, we select the `"interval"` method when calling the function, and it automatically selects non-overlapping sections of the data for regressions (**Fig. 2b**).

```{r Figure2, echo = F, fig.cap="Illustration showing the sampling method used by `auto.rate()` during **(A)** interval regression analysis, where sample windows do not overlap and **(B)** rolling regression analysis, where sample windows overlap and move forward by one sample unit at a time."}

@@

To estimate the most linear sections of the data, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) rate. If the rate changes at some point during the sampling period, then the rolling estimate should detect the variability. This function takes advantage of this relationship and performs a Gaussian kernel density estimate (KDE) of all regressions [see @silverman_density_1986]. Peaks in the density thus reflect stable rates, and new segments of the data within the kernel bandwidth of the peak values are resampled and re-analysed to determine the most linear sections. Summary diagnostics and visual plots provide supporting information on the technique and the validity of the results (**Fig. 3**).

The function `calc.rate.ft()` is used to specifically analyse flow-through respirometry data, and is similar to `calc.rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate.`calc.rate.ft()` can process data to determine the mean or modal $RO_2$ over a period.

```{r Figure3, echo = F, fig.cap="Typical graphical outputs of the function `auto.rate()`. **(A)** Depending on the `method` input, the function is able to show: (clockwise form top left) the maximum rate detected, the minimum rate detected, calculation of rate by fixed-width intervals, and the most linear region as determined by kernel density estimates. **(B)** In addition, the outputs will also print diagnostic plots of: (clockwise from top left) the rolling regression of rate as a function of time with dotted lines indicating the region used to estimate the parameter, a kernel density plot that shows only when the `linear` method is called, a residual plot of the regression, and a normal Q-Q plot of the residuals of the regression. Analysis is performed using the example dataset `sardine.rd`, which is included in the package. Code to reproduce the plots is available in the Supplementary Material."}

@@

The function `adjust.rate()` can be used to correct $RO_2$ estimates for oxygen flux and/or background respirometry data. As the oxygen consumption units are not mass specific at the point of correction, the function adjusts RO_2 directly without bias using the equation: $$RO_{2(correction)} = RO_2 - (\phi_d + RO_{2(bg)})$$ where $\phi_d$ is the oxygen flux and $RO_{2(bg)}$ is background respiration.

@@

For deliberate reasons, units of measurement are not used when determining $RO_2$ in `respR`. Instead, two functions, `convert.rate()` and `scale.rate()`, are used to transform $RO_2$ to specific units and perform volumetric or mass-specific conversions . This unique approach has the advantage of allowing "dimensionless" rate calculations for data that may not be fully supported by the entire respR workflow. For example, measurements of respiration can be scaled to surface area (e.g. diatoms, Iversen & Ploug, 2013, Naumann et al., 2014), volume (e.g. copepod eggs, Hammervold et al., 2015), or density (e.g. sea urchin larvae, Stumpp et al., 2011), all of which are currently not supported in `respR`. Users may also be interested in using respR's analytical functions to investigate other linear relationships or perform breakpoint analyses on non-respirometry-related data.

The function `convert.rate()` is used to convert units of oxygen concentration only (e.g. $mg/L$ or %) and has support for various unit combinations (Table 2). For ease of use, we have implemented a simple fuzzy string-matching algorithm to match relevant units even when they have been provided in different formats (e.g. "$mg/L$", "$mg/l$", "$mg L^{-1}$" and "$mg l^{-1}$" are identified as the same units). The function `scale.rate()` is conceptually similar to `convert.rate()`, but it focuses volumetric and mass-specific conversions (e.g. $mg h^{-1} kg^{-1}$).

@@

>>>>

4. Although designed specifically with aquatic respirometry in mind, the object-oriented approach of the package and the dimensionless nature of its analytical functions mean that parts of the package can easily be used to estimate linear relationships from a range of applications in many research disciplines."

keywords: "R, software, respirometry, ecophysiology, metabolic rate"
thanks: "Replication files are available on the first author's Github account (http://github.com/januarharianto). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: nicholas.carey@sams.ac.uk."

@@

opts_chunk$set(collapse = TRUE, comment = "#>", tidy = TRUE, cache = TRUE, 
  highlight = TRUE)
```

# Introduction
Metabolic rate is a fundamental trait associated with virtually all biological functions and is key in predicting patterns in ecology and conservation biology, from populations [@seibel_rate_2007; @barneche_scaling_2014] to ecosystems [@brown_toward_2004]. It has been widely investigated in studies of the effects of external stressors on organisms, and is increasingly used to study resilience in climate change-related studies where warming is expected to drive increased metabolism in ectotherms [e.g. @portner_climate_2002; @carey_sea_2016; @delorme_effects_2016]. By far the most common metric used to assess physiological performance in animals, metabolic rate has been determined from whole organisms to the level of cells and tissues [@white_determinants_2013]. Metabolic rate is typically quantified using respirometry, which measures the rate of oxygen uptake over time, $RO_2$ [@lighton_measuring_2008].

<!-- There are four broad methodological approaches in respirometry: closed, intermittent flow, flow-through and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow mixing of the environment within the chamber. Oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from the $O_2$ timeseries by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$VO_2 = \Delta O_2V$$ where $\Delta O_2$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container [@lighton_measuring_2008].  -->

<!-- In intermittent flow respirometry, $O_2$ concentration is measured as described above, but periodically the chamber is flushed with new water or air, returning it to initial conditions, resealed, and the experiment repeated [@svendsen_design_2016]. This technique is essentially closed respirometry, but with the incorporation of repeated measures. Depending on the metabolic rate metric being investigated, final respiration rate can be calculated as the mean of the measures [e.g. @carey_sea_2016], or the lowest or highest rates recorded in any trial [e.g. @stoffels_physiological_2015]. -->

<!-- Flow-through respirometry involves a closed chamber, but with a regulated flow of air or water through it at a precisely determined rate. After equilibrium has been achieved, the oxygen concentration differential between the incurrent and excurrent channels, along with the flow rate, allows calculation of the oxygen extracted from the flow volume per unit time: $$VO_2 = (C_iO_2 - C_eO_2)FR$$ where $VO_2$ is the rate of $O_2$ consumption over time, $C_iO_2$ and $C_eO_2$ are the incurrent and excurrent $O_2$ concentrations, and $FR$ is the flow rate of water through the system [@lighton_measuring_2008]. -->

<!-- A final method is open respirometry, in which an open tank or semi-enclosed area is used, but the input or mixing rate of oxygen from the surroundings is known or found to be negligible relative to oxygen consumption of the specimens [@leclercq_measurement_1999]. It is seldom used, but for some applications it is a sufficient and practical methodology [@gamble_open-top_2014]. The common equation used for open respirometry is: $$VO_2 = \Delta O_2V + \phi_d$$ where $\Delta O_2V$ is the slope of the regression that relates $O_2$ concentration to time, $V$ is the volume of the container and $\phi_d$ is the oxygen flux as determined by Fick's Law [@leclercq_measurement_1999]. -->

@@

Respirometry studies are increasingly outputting large, high-resolution datasets  conducted over longer periods [e.g. 20 h ; @norin_intraspecific_2012]. In most cases, processing the data involves an *ad hoc* selection of data points with poorly reported criteria, and subsequent manual processing of the data subset(s) using a spreadsheet program (e.g. Microsoft Excel) or an integrated development environment (IDE, e.g. R or Matlab). These approaches can be tedious and time consuming especially when spreadsheet programs struggle with the vast datasets that are generated, while IDEs require a degree of expertise to use and have substantial learning curves. Dedicated software are also available to perform metabolic rate analyses, but many have costly licensing requirements (e.g. AutoResp by Loligo Systems), complicating or preventing their use on multiple machines, and are proprietary and closed-source, hindering scientific reproducibility and transparency.

A number of open-source R software packages have recently become available which are designed for, or are suitable to analyse respirometry data. The "`respirometry`"" package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The "`rMR`"" package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). The package "`LoLinR`"" [@olito_estimating_2017], while not specifically coded for respirometry, provides a statistically robust method of detecting a "best-fit" regression, and performs well at identifying truly linear subsets of a data series. Importantly, all of these packages assume that data are monotonic with evenly-spaced sampling periods with respect to time, an ideal structure but one where respirometry data often fails to meet. They also do not have the tools to analyse complicated datasets to automatically determine useful metric such as $MO_{2,max}$ and $MO_{2,min}$.

Here we describe `respR`, an open-source R package designed to provide an efficient and reproducible workflow for the analysis of respirometry data (Table 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data [see @lighton_measuring_2008; @svendsen_design_2016], (2) determine volume, mass, and surface area specific oxygen uptake rates, (3) automatically detect maximal and minimal rates, $P_{crit}$, and (4) detect the most linear sections of data using traditional rolling regression and kernel density estimation techniques. The package facilitates the output as supplementary material of all stages of respirometry analyses; raw data, subsetting criteria, background corrections, units of measure, and resulting metabolic rates. This streamlines the reporting of analytical methods and allows reviewers or other investigators to fully reproduce or scrutinise the results. Other functions useful in processing respirometry data are avaliable and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). The utility of the `respR` package is shown here using  data collected for closed, intermittent and flow-through respirometry. We show how `respR` might be used to determine $MO_{2,max}$, $MO_{2,min}$, and $P_{crit}$.

@@

The `respR` package streamlines the analysis of respirometry data by partitioning the data processing workflow into several independent actions: (1) initially, data are imported, checked for common errors, plotted for visual inspection and exported as an object; (2) the object is then analysed to estimate the desired rate parameter; (3) diagnostic plots are automatically generated and the data can be further explored using the generic S3 `print()`, `plot()`, and `summary()` commands to verify the results; (4) adjustments to the data, either from background respiration and/or oxygen flux in open respirometry, are applied; (5) volume or mass-specific rate conversions, specific to the respiratory chamber and specimen(s), are calculated where necessary (Fig. 1). By design, each step of the workflow requires minimal manual manipulation of the data, and the user can pass the output of each function to the next. This object-oriented approach makes `respR` easy to use, even for novice users of R, and allows any interested R programer to easily extend the functionality of the package, either by adding additional intermediate actions or substituting current actions with other code without breaking the rest of the package. Note, we use $RO_2$ in the following text to represent the dimensionless rate of $O_2$ use with regard to the time unit, and to distinguish it from other commonly reported metrics (e.g. $MO_2$, $VO_2$).

@@

Data should be formatted correctly before use in `respR`. The function `inspect_data()` extracts a two-column data frame from a multi-column dataset and performs simple error checks while plotting the data for quick data visualisation. All data should be numeric; `respR` at this time has limited support for date/time data (e.g. `POSIXct` and `POSIXt` classes). Parsing  different time formats is extremely complex and beyond the scope of this package, particularly because such packages already exist (e.g. `lubridate`). Time data, being a continuous variable, should be sequential, without duplicates. Ideally, sampling frequency will be monotonic, but in many cases this may not be the case, such as experiments with irregular sampling intervals, or because machine drift and/or data dropouts frequently occur during regular sampling. Therefore, `respR` uses time-based data extraction methods by default in most of its analytical functions (except `pcrit()`), and correctly handles irregular timeseries data.

Using `inspect_data()` is optional -- the main functions in our package will readily work with data frames that were not processed by the function. Running `inspect_data()` is an exploratory step that flags potential issues before analysis. Error check results are summarised in the R console and all tests are saved in the output object with the locations (i.e. row numbers) of the errors:

@@

Background oxygen consumption rate, $RO_{2(bg)}$, is often measured to account for the contribution of microbial respiration, and applied as a correction to experimental rates [@rogers_new_2016]. Background rates typically account for a small percentage of experimental rates, and multiple “blank” experiments are usually conducted and the rates averaged to obtain a more accurate estimate of the correction [e.g. @carey_size_2014; @daoud_temperature_2007]. The function `calc_rate.bg()` uses simple linear regression to simultaneously process multiple background rate measurements. Data segments can be truncated before analysis by time period or row numbers. The results are stored as an object for later use.

@@

Typically, the analysis of respirometry data requires the selection one or more linear sections of the data for more accurate calculations of $RO_2$ [e.g. @gordon_energetics_1989; @chu_ecophysiological_2017; @carey_sea_2016]. The function `calc_rate()` can manually extract and process data subset from respiromety data. Data subsets can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accommodate most, if not all, manual data selection requirements and allow for consistent reporting of methods and results. Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results. Generally, $RO_2$ is calculated using the equation: $$RO_2 = \Delta O_2V$$ where $\Delta O_2$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container [@lighton_measuring_2008].

### Estimating maximum, minimum, most linear and interval $RO_2$

For more complex respirometry parameters, the function `auto_rate()` uses novel rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs regressions of sample size $n$ iteratively across the length of the data (i.e. rolling regressions), one sample at a time (Fig. 2a). Thus, a total of $(m - n) + 1$ number of overlapping regressions are fitted, which are then ranked to obtain maximum and minimum values. If an interval-based regression is required, we select the `"interval"` method when calling the function, and it automatically selects non-overlapping sections of the data for regressions (Fig. 2b).

```{r Figure2, echo = F, fig.cap="Illustration showing the sampling method used by `auto_rate()` during **(A)** interval regression analysis, where sample windows do not overlap and **(B)** rolling regression analysis, where sample windows overlap and move forward by one sample unit at a time."}

@@

To estimate the most linear sections of the data, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) rate. If the rate changes at some point during the sampling period, then the rolling estimate should detect the variability. This function takes advantage of this relationship and performs a Gaussian kernel density estimate (KDE) of all regressions [see @silverman_density_1986]. Peaks in the density thus reflect stable rates, and new segments of the data within the kernel bandwidth of the peak values are resampled and re-analysed to determine the most linear sections. Summary diagnostics and visual plots provide supporting information on the technique and the validity of the results (Fig. 3).

### Calcularing $RO_2$ from flow-through respirometry

The function `calc_rate.ft()` is used to specifically analyse flow-through respirometry data, and is similar to `calc_rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate, based on the equation: $$VO_2 = (C_iO_2 - C_eO_2)FR$$ where $VO_2$ is the rate of $O_2$ consumption over time, $C_iO_2$ and $C_eO_2$ are the incurrent and excurrent $O_2$ concentrations, and $FR$ is the flow rate of water through the system [@lighton_measuring_2008].

```{r Figure3, echo = F, fig.cap="Typical graphical outputs of the function `auto_rate()`. **(A)** Depending on the `method` input, the function is able to show: (clockwise form top left) the maximum rate detected, the minimum rate detected, calculation of rate by fixed-width intervals, and the most linear region as determined by kernel density estimates. **(B)** In addition, the outputs will also print diagnostic plots of: (clockwise from top left) the rolling regression of rate as a function of time with dotted lines indicating the region used to estimate the parameter, a kernel density plot that shows only when the `linear` method is called, a residual plot of the regression, and a normal Q-Q plot of the residuals of the regression. Analysis is performed using the example dataset `sardine.rd`, which is included in the package. Code to reproduce the plots is available in the Supplementary Material."}

@@

The function `adjust.rate()` can be used to correct $RO_2$ estimates for oxygen flux and/or background respirometry data. As the oxygen consumption units are not mass specific at the point of correction, the function adjusts RO_2 directly without bias using the equation: $$RO_{2(correction)} = RO_2 - (\phi_d + RO_{2(bg)})$$ where $\phi_d$ is the oxygen flux as determined by Fick's Law [@leclercq_measurement_1999] and $RO_{2(bg)}$ is background respiration.

@@

Units of measurement are not used when determining $RO_2$ in `respR`. Instead, two functions, `convert_DO()` and `convert_rate()`, are used to transform $RO_2$ to specific units and perform volumetric or mass-specific conversions . This unique approach has the advantage of allowing "dimensionless" rate calculations for data that may not be fully supported by the entire respR workflow. For example, measurements of respiration can be scaled to surface area [e.g. diatoms & coral, @iversen_temperature_2013; @naumann_species-specific_2014], volume [e.g. copepod eggs, @hammervold_new_2015], or density [e.g. sea urchin larvae, @stumpp_co2_2011], all of which are currently not supported in `respR`. Users may also be interested in using respR's analytical functions to investigate other linear relationships or perform breakpoint analyses on non-respirometry-related data.

The function `convert_DO()` is used to convert units of oxygen concentration only (e.g. $mg/L$ or %) and has support for various unit combinations (Table 2). For ease of use, we have implemented a simple fuzzy string-matching algorithm to match relevant units even when they have been provided in different formats (e.g. "$mg/L$", "$mg/l$", "$mg\ L^{-1}$" and "$mg\ l^{-1}$" are identified as the same units). The function `convert_rate()` is conceptually similar to `convert_DO()`, but it focuses volumetric and mass-specific conversions (e.g. $mg\ h^{-1}\ kg^{-1}$).

@@

# Acknowledgements
This work was supported by an NSW Environmental Trust grant (MB), and Endeavour Postdoctoral Fellowship (NC) and a Professor NGW & Mrs Ann Macintosh Memorial Scholarship from the University of Sydney (JH) and is Sydney Institue of Marine Science contribution number XXXX (to be confirmed).

# Author contributions
JH, NC and MB conceived the ideas and designed methodology; JH and NC wrote the code and analysed the example data; JH, NC and MB led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.
